---
title: "Courier Newsroom Spend"
author: "CJ Robinson"
date: "2024-11-03"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(ggthemes)
library(gt)
library(RColorBrewer)

```

# Read in data

```{r}
meta_ads_raw <- read_csv("../data/consolidated_ads_20241103_173021.csv") %>% clean_names()
```


# Visualizations

# --------

# Courier comparisons

```{r}
meta_ads_raw %>% 
  filter(year(ad_creation_time) %in% c(2020, 2022, 2024)) %>%
  # Categorize newsrooms based on page names, starting with courier since they have a much smaller amount of newsrooms.
  mutate(network = case_when(
    page_name %in% c('The Keystone', 'The \'Gander Newsroom', 'UpNorthNews', 'Cardinal & Pine', 'The Copper Courier', 'Iowa Starting Line', "The Nevadan", "Granite Post") ~ "Courier Newsroom",
    page_name == "The Morning Mirror" ~ "Star Spangled Media",
    page_name == "Olean Star" ~ "Star News",
    page_name %in% c('The Michigan Independent', 'The Pennsylvania Independent', 'The Wisconsin Independent', 'The Nebraska Independent', 'The Montana Independent') ~ "American Independent",
    # Anything not above is considered metric media
    .default = "Metric Media"
  )) %>%
  mutate(courier_flag = ifelse(network == "Courier Newsroom", "Courier Newsroom", "Other"),
         year = year(ad_creation_time)) %>%
  # Group by week and network
  group_by(network, courier_flag, year) %>%
  # Sum spend (we'll use low spend as a minimum)
  summarize(upper_sum = sum(high_spend),
            low_sum = sum(low_spend, na.rm = TRUE),
            avg_sum = (sum(high_spend) + sum(low_spend))/2) %>%
  ungroup() %>%
  complete(year = c(2020, 2022, 2024), network, fill = list(low_sum = 0, upper_sum = 0, avg_sum = 0)) %>%
  ggplot(aes(x = factor(year), y = low_sum, fill = reorder(network, desc(low_sum)))) +
  geom_bar(stat = "identity", position = position_dodge(width = 1), width = 1) +
  scale_y_continuous(label = scales::dollar_format(scale = .000001, suffix = "M")) +
  labs(title = "Courier Massively Outspent Other Networks", 
       subtitle = "Total spending on Meta ads by network",
       fill = "Network",
       x = "Year") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.line.x = element_line(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    plot.title = element_text(size = 20),
    plot.subtitle = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14)
  ) +
  scale_fill_manual(values = c(
    "Courier Newsroom" = "#E9381F",
    "Star Spangled Media" = "#757575",
    "Metric Media" = "#9E9E9E",
    "Star News" = "#BDBDBD",
    "American Independent" = "#E5E4E2"
  ))

ggsave("../img/courier_comp.png", height = 6, width = 10)


```

# Chart 2 - Meta table

```{r}
meta_ad_report <- read_csv("../data/FacebookAdLibraryReport_2024-10-30_US_last_30_days_advertisers.csv") %>% clean_names()

table <- meta_ad_report %>% 
  mutate(categorize_page = case_when(
    disclaimer %in% c("HARRIS FOR PRESIDENT", "HARRIS VICTORY FUND") ~ "Harris For President or Harris Victory Fund",
    page_name %in% c('The Keystone', 'The \'Gander Newsroom', 'UpNorthNews', 'Cardinal & Pine', 'The Copper Courier', 'Iowa Starting Line', "The Nevadan", "Granite Post") ~ "Courier Newsroom", 
    .default = page_name
  )) %>% 
  dplyr::group_by(categorize_page) %>% 
  summarize(amount_spent_usd = sum(as.numeric(amount_spent_usd))) %>% 
  arrange(desc(as.numeric(amount_spent_usd))) %>% 
  top_n(10)

table_viz <- table %>% 
  dplyr::select(categorize_page, amount_spent_usd) %>% 
  gt() %>% 
  tab_header(
    title = "When combined, Courier was the third highest spender on Meta Ads",
    subtitle = "Top Meta ad spenders from October 1 to October 30, 2024"
  ) %>% 
  fmt_number(decimals = 2) %>% 
  fmt_currency(columns = amount_spent_usd, decimals = 0) %>% 
  cols_label(categorize_page = "Spender", amount_spent_usd = "Amount Spent (USD)")  %>%  
  text_transform(
    locations = cells_body(
      rows = 3 # Specify which row to bold (in this case, row 2)
    ),
    fn = function(x) paste0("<b>", x, "</b>"))

table_viz

gtsave(table_viz, "../img/table.png")
```


# Chart 3 - Targetings

```{r}
library(jsonlite)

expand_demographic_data <- function(demographic_list) {
  # Check if demographic_list is not NA or empty
  if (is.na(demographic_list) || demographic_list == "") {
    return(list())  # Return an empty list if the data is missing
  }
  
  # Convert string to JSON
  demographic_list <- fromJSON(gsub("'", '"', demographic_list), simplifyVector = FALSE)
  
  demographic_data <- list()
  for (entry in demographic_list) {
    if (is.list(entry) && !is.null(entry$gender) && !is.null(entry$age) && !is.null(entry$percentage)) {
      column_name <- paste(entry$gender, entry$age, sep = "-")
      demographic_data[[column_name]] <- as.numeric(entry$percentage)
    }
  }
  return(demographic_data)
}

# Apply the function across all rows in 'demographic_distribution'
expanded_demographics <- lapply(meta_ads_raw$demographic_distribution, expand_demographic_data)

# Collect all unique column names
all_columns <- unique(unlist(lapply(expanded_demographics, names)))

# Convert expanded_demographics into a data frame with all columns, filling missing ones with 0
expanded_df <- bind_rows(lapply(expanded_demographics, function(row) {
  row[setdiff(all_columns, names(row))] <- 0  # Add missing columns with 0
  row
}))

# Restore the original row index
rownames(expanded_df) <- rownames(meta_ads_raw)

# Combine with the original data frame
meta_ads_expanded <- cbind(meta_ads_raw, expanded_df)

# Verify the result
print(paste("Original rows:", nrow(meta_ads_raw)))
print(paste("Expanded rows:", nrow(meta_ads_expanded)))
meta_ads_expanded <- meta_ads_expanded  %>%  mutate(
    male_total = rowSums(across(starts_with("male"))),
    female_total = rowSums(across(starts_with("female")))
  )

meta_ads_expanded %>% 
    filter(ad_delivery_start_time >= '2024-01-01', 
           #make sure null demographic info is taken out
           male_total + female_total > 0) %>% 
  # Categorize newsrooms based on page names, starting with courier since they have a much smaller amount of newsrooms.
  mutate(network = case_when(
    page_name %in% c('The Keystone', 'The \'Gander Newsroom', 'UpNorthNews', 'Cardinal & Pine', 'The Copper Courier', 'Iowa Starting Line', "The Nevadan", "Granite Post") ~ "Courier Newsroom",
    page_name == "The Morning Mirror" ~ "Star Spangled Media",
    page_name == "Olean Star" ~ "Star News",
    page_name %in% c('The Michigan Independent', 'The Pennsylvania Independent', 'The Wisconsin Independent', 'The Nebraska Independent', 'The Montana Independent') ~ "American Independent",
    # Anything not above is considered metric media
    .default = "Metric Media"
  )) %>%
  filter(page_name == "The Keystone") %>% 
  ggplot(aes(x = female_total)) + 
  geom_histogram(bins =30, fill = "#08519C")+
  geom_vline(aes(xintercept = .5), linetype="dotted") +
  labs(title = "The Keystone's Meta ads reached more women than men", 
       subtitle = "Ad count for Courier's newsroom by gender reach") +
  xlab("Percentage of Women Reached of Total Population") +
  theme_minimal() +
  scale_x_continuous(labels = scales::percent_format()) +
  theme(axis.title.y=element_blank(),
        legend.position="none",
        axis.title.x = element_text(size = 14),
        axis.line.x = element_line(),
        panel.grid.major.x=element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 16)) + 
  annotate("text", x = 0.34, y = 35, label = "Most ads reached more women", size = 6, hjust = 1) +
  annotate("curve", x = 0.35, y = 35, xend = 0.55, yend = 35, curvature = - 0.25, arrow = arrow(length = unit(0.5, "cm"))) +
  annotate("text", x = 0.43, y = 17, label = "Some targeted only men", size = 6, hjust = 1) +
  annotate("curve", x = 0.3, y = 15, xend = 0.03, yend = 6, curvature = -0.25, arrow = arrow(length = unit(0.5, "cm"))) 

ggsave("../img/keystone_dist.png", height = 6, width = 10)
```


# Chart 4 - Monthly by Page Breakout for Courier

```{r}
meta_ads_raw %>% 
    filter(ad_delivery_start_time >= '2024-08-01') %>% 
  # Categorize newsrooms based on page names, starting with courier since they have a much smaller amount of newsrooms.
  mutate(network = case_when(
    page_name %in% c('The Keystone', 'The \'Gander Newsroom', 'UpNorthNews', 'Cardinal & Pine', 'The Copper Courier', 'Iowa Starting Line', "The Nevadan", "Granite Post") ~ "Courier Newsroom",
    page_name == "The Morning Mirror" ~ "Star Spangled Media",
    page_name == "Olean Star" ~ "Star News",
    page_name %in% c('The Michigan Independent', 'The Pennsylvania Independent', 'The Wisconsin Independent', 'The Nebraska Independent', 'The Montana Independent') ~ "American Independent",
    # Anything not above is considered metric media
    .default = "Metric Media"
  )) %>%
  filter(network == "Courier Newsroom") %>% 
  #group by week and network
  group_by(week = floor_date(ymd(ad_delivery_start_time), unit = "week"), page_name) %>% 
  #sum spend (we'll use low spend as a minimum)
  summarize(low_sum = sum(low_spend, na.rm = TRUE)) %>% 
  ggplot(aes(x = week, y = low_sum, fill = reorder(page_name, (low_sum)))) + 
  geom_bar(stat = "identity", color = "white") +
  scale_y_continuous(label = scales::dollar_format(scale = .000001, suffix = "M", accuracy = .1)) +
  labs(title = "Courier Concentrated Spending in Pennsylvania and North Carolina", 
       subtitle = "Weekly Meta ad spend by Courier, 2024" ) +
  scale_fill_brewer(palette = "Blues",
                    name = "Courier Outlet", 
                      labels=c("Iowa Starting Line" = "Iowa - Iowa Starting Line",
                            "Granite Post" = "New Hampshire - Granite Post",
                            "UpNorthNews" = "Wisconsin - UpNorthNews",
                            "The Nevadan" = "Nevada - The Nevadan",
                            "The Copper Courier" = "Arizona - The Copper Courier",
                            "The 'Gander Newsroom" = "Michigan - The 'Gander",
                            "Cardinal & Pine" = "North Carolina - Cardinal & Pine",
                            "The Keystone" = "Pennsylvania - The Keystone")) +
  theme_minimal() +
  theme(axis.title=element_blank(),
        axis.line.x = element_blank(),
        panel.grid.major.x=element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14)) 

ggsave("../img/courier_state_spend.png", height = 6, width = 10)

```


# Other analysis

# -----

# Export data for inspection of GOTV

Quote: 35 of the 48 Meta ads in October which cost at least $30,000 each directed users to voting resources, often linking to guides included in each newsroom’s website. These ads cost the newsrooms at least $1.5 million in total. 

Pulled from body and facebook ad library manually to code each of the top 48 ads in Excel sheet linked included in Github

```{r}
meta_ads_expanded %>% 
    filter(ad_delivery_start_time >= '2024-10-01',
           low_spend >= 10000) %>% 
    arrange(desc(low_spend)) %>% 
  write_csv("top_oct_ad_spends.csv")
```


# Looking into currently running ads

Quote: Before the final cut-off date for launching political ads, Courier spent around $2.4 million on over 400 ads that will continue to run during the election. 

```{r}
meta_ads_expanded %>% 
  filter(ad_delivery_start_time >= '2024-01-01',
         page_name %in% c('The Keystone', 'The \'Gander Newsroom', 'UpNorthNews', 'Cardinal & Pine', 'The Copper Courier', 'Iowa Starting Line', "The Nevadan", "Granite Post"),
         is.na(ad_delivery_end_time) | ad_delivery_end_time > "2024-11-03") %>% 
  summarize(low_spend = sum(low_spend), 
            count = n())
```

# What was the impact of men-only ads? 

Quote: 16 ads were targeting exclusively toward women

```{r}
meta_ads_expanded %>% 
  filter(ad_delivery_start_time >= '2024-10-01',
         page_name == "The Keystone",
         female_total + male_total > 0) %>% 
  group_by(male_only = ifelse(female_total==0, "male only","other"))%>% 
  summarize(count = n(),
            low_spend = sum(low_spend, na.rm = TRUE), 
            high_spend = sum(high_spend, na.rm = TRUE))
```

# Relentless


```{r}
meta_ads_expanded %>% 
    filter(ad_delivery_start_time >= '2024-01-01',
           grepl("Relentless", ad_creative_bodies)) %>% 
  #categorizes newsrooms based on page names, starting with courier since they have a much smaller amt of newsrooms. 
  filter(page_name %in% c('The Keystone', 'The \'Gander Newsroom', 'UpNorthNews', 'Cardinal & Pine', 'The Copper Courier', 'Iowa Starting Line', "The Nevadan", "Granite Post")) %>% 
  summarize(count = n(),
            low_spend = sum(low_spend, na.rm = TRUE))
```

# -----

# Alternative targeting image

```{r}
meta_ads_expanded %>% 
    filter(ad_delivery_start_time >= '2024-10-01') %>% 
  #categorizes newsrooms based on page names, starting with courier since they have a much smaller amt of newsrooms. 
  mutate(network = case_when(
    page_name %in% c('The Keystone', 'The \'Gander Newsroom', 'UpNorthNews', 'Cardinal & Pine', 'The Copper Courier', 'Iowa Starting Line', "The Nevadan", "Granite Post") ~ "Courier Newsroom", 
    page_name == "The Morning Mirror" ~ "Star Spangled Media",
    #anything not above is considered metric media
    .default = "Metric Media")) %>% 
  filter(page_name == "The Keystone") %>% 
  mutate(female_total = case_when(
    female_total <= 0.1 ~ 1,
    female_total <= 0.2 ~ 2,
    female_total <= 0.3 ~ 3,
    female_total <= 0.4 ~ 4,
    female_total <= 0.5 ~ 5,
    female_total <= 0.6 ~ 6,
    female_total <= 0.7 ~ 7,
    female_total <= 0.8 ~ 8,
    female_total <= 0.9 ~ 9,
    TRUE ~ 10
  ))  %>% 
  group_by(female_total) %>% 
  summarize(low_spend = sum(low_spend, na.rm = TRUE)) %>% 
  ggplot(aes(x = female_total / 10, y = low_spend)) + 
  geom_bar(stat = "identity",  fill = "#08519C") + 
  # geom_vline(aes(xintercept = .5), linetype="dotted") +
  labs(title = "Courier  Pennsylvania outlet's reached ads more toward women", 
       subtitle = "The Keystone 2024 Meta ad spend by percentage targeting women") +
  xlab("Percentage of Women Targeted") +
  theme_minimal() +
   scale_x_continuous(labels = scales::percent_format()) +
   scale_y_continuous(labels = scales::label_dollar()) +
  theme(axis.title.y=element_blank(),
        legend.position="none",
        axis.title.x = element_text(size = 14),
        axis.line.x = element_line(),
        panel.grid.major.x=element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 16)) 
  # annotate("text", x = 0.34, y = .13, label = "Most ads targeted women", size = 6, hjust = 1) +
  # annotate("curve", x = 0.35, y = .13, xend = 0.55, yend = .15, curvature = - 0.25, arrow = arrow(length = unit(0.5, "cm"))) +
  # annotate("text", x = 0.43, y = .07, label = "Some targeted only men", size = 6, hjust = 1) +
  # annotate("curve", x = 0.3, y = .056, xend = 0.03, yend = .03, curvature = -0.25, arrow = arrow(length = unit(0.5, "cm"))) 


ggsave("../img/keystone_dist_alt.png", height = 6, width = 10)
```